Clinical advancement forecasting \thanks{\textit{\underline{Citation}}: \textbf{TBD}}

### 2024-03-30 paper/paper.tex
Minor figure and caption improvements

Figure~\ref{fig:performance_across_ta} demonstrates how well our primary model in this study -- a constrained, linear L2-regularized model referred to from here on as "RDG" -- ranks TD pairs by comparison to a composite score from Open Targets ("OTS"). The relative risk (RR) metric used in this comparison as well as all others where not specified otherwise is defined as:

@@

The third ranking method presented in Figure~\ref{fig:performance_across_ta}, "RDG-T", differs from the RDG model only in that it also uses time since the phase 2 transition as a predictive factor as well as all others. We observe that the use of this information almost universally improves standard performance metrics like ROC (receiver operating characteristic) and AP (average precision), however it adds little to no value in rankings beyond a level where substantial relative risk increases can be observed. In other words, it constitutes an effective but coarse mechanism for ranking TD pairs while lacking the high precision of other factors like genetic support. This is consistent with its nature as a necessary but not sufficient condition for success, and it is likely that much of the value it adds in the tail of lower rankings could be captured and enhanced if other early indicators for the many reasons trials fail \cite{Razuvayevskaya2023.02.07.23285407} were also included (see more on this in Section~\ref{sec:discussion}). As a more practical concern, we refrain from focusing on RDG-T, or the similar GBM-T model, because neither is readily applicable to undeveloped TD pairs for which the time since phase 2 transition is not available. They are, however, a useful performance ceiling towards which future work might build towards.

A subset of therapeutic areas was used for this analysis and this is reflected in both the mean estimates and distributions of Figure~\ref{fig:performance_across_ta}. A full list of these can be seen in Supplementary~Figure~\ref{fig:relative_risk_by_ta}, which also shows the RR estimates used in any summary statistics across therapeutic areas. These were selected based on criteria described in Section~\ref{sec:methods}. Lastly, Supplementary~Figure~\ref{fig:relative_risk_dist_across_ta} shows RR distributions by model with significances of their differences.

@@

@@

@@

\begin{figure}[!htb]
  \centering
  \captionsetup{width=.9\linewidth}
  \includegraphics[width=1\textwidth]{opportunity_summary.png}
  \caption{
    \textbf{Present-day target-disease pair counts by tractability, likelihood of advancement and therapeutic area}. The \textbf{stage} panel contains counts by maximum trial phase reached, the \textbf{threshold} panel contains counts of pairs with a RDG model score exceeding that of the associated benchmark for only undeveloped pairs, and the \textbf{tractability} panel shows pair frequencies among undeveloped pairs exceeding the EVA threshold that also have a HIGH tractability rating as defined in Supplementary~Table~\ref{tab:tractability_buckets}. This corresponds to targets that have all been in clinical development already, except for the \textbf{OC} modality in which case it indicates that a target has been approved. 
  }
  \label{fig:opportunity_summary}
\end{figure}

@@

\subsection{Algorithms}

The primary model for this analysis, RDG, is a constrained, L2-regularized linear regressor. Alternatives to this choice we explore include constrained and unconstrained gradient-boosted machine (GBM) models fit with the LightGBM \cite{LightGBM} algorithm as well as unconstrained L2-regularized linear regressors (i.e. RDG with no constraints). The performance of these alternatives, as quantified by relative risk estimates among top predictions, is illustrated in Figure~\ref{fig:relative_risk_model_features}.

\begin{figure}[!htb]
  \centering
  \captionsetup{width=.9\linewidth}
  \includegraphics[width=1\textwidth]{relative_risk_model_features.pdf}
  \caption{
    \textbf{Performance by algorithm, constraint type and feature grouping}.  
  }
  \label{fig:relative_risk_model_features}
\end{figure}

@@

\begin{figure}
  \centering
  \captionsetup{width=.9\linewidth}
  \includegraphics[width=1\textwidth]{relative_risk_core_features.pdf}
  \caption{
    \textbf{Performance of individual features and predictive scores as measured by relative risk}.
    RDG model results denoted by \colorbox{Gainsboro}{rdg@N} correspond to results for the \colorbox{Gainsboro}{N} top TD pairs. The same convention is used for \colorbox{Gainsboro}{literature} evidence and the \colorbox{Gainsboro}{time\_\_transition\_\_eq\_X} convention denotes RR estimates when the time since the phase 2 transition is equal to \colorbox{Gainsboro}{X} years. All other features are assessed based on their existence. The counts along the origin indicate how many TD pairs were used to compute the RR numerator.
  }
  \label{fig:relative_risk_core_features}
\end{figure}

@@

>>>>

Figure~\ref{fig:performance_across_ta} demonstrates how well our primary model in this study -- a constrained, linear L2-regularized model referred to from here on as "RDG" (short for ridge regression) -- ranks TD pairs by comparison to a composite score from Open Targets ("OTS"). The relative risk (RR) metric used in this comparison as well as all others where not specified otherwise is defined as:

@@

The third ranking method presented in Figure~\ref{fig:performance_across_ta}, "RDG-T", differs from the RDG model only in that it also uses time since the phase 2 transition as a predictive factor as well as all others. We observe that the use of this information almost universally improves standard performance metrics like ROC (receiver operating characteristic) and AP (average precision), however it adds little to no value in rankings beyond a level where substantial relative risk increases can be observed. In other words, it constitutes an effective but coarse mechanism for ranking TD pairs while lacking the high precision of other factors like genetic support. This is consistent with its nature as a necessary but not sufficient condition for success, and it is likely that much of the value it adds in the tail of lower rankings could be captured and enhanced if other early indicators for the many reasons trials fail \cite{Razuvayevskaya2023.02.07.23285407} were also included (see more on this in Section~\ref{sec:discussion}). As a more practical concern, we refrain from focusing on RDG-T, or the similar GBM-T model, because neither is readily applicable to undeveloped TD pairs for which the time since phase 2 transition is not available. They are, however, a useful performance ceiling towards which future work could build.

A subset of therapeutic areas was used for this analysis and this is reflected in both the mean estimates and distributions of Figure~\ref{fig:performance_across_ta}. A full list of these can be seen in Supplementary~Figure~\ref{fig:relative_risk_by_ta}, which also shows the RR estimates used in any summary statistics across therapeutic areas. These were selected based on criteria described in Section~\ref{sec:methods}. Lastly, Supplementary~Figure~\ref{fig:relative_risk_dist_across_ta} shows RR distributions by model with the significance of their differences.

@@

\begin{figure}[!htb]
  \centering
  \captionsetup{width=.9\linewidth}
  \includegraphics[width=1\textwidth]{relative_risk_core_features.pdf}
  \caption{
    \textbf{Performance of individual features and predictive scores as measured by relative risk}.
    RDG model results denoted by \colorbox{Gainsboro}{rdg@N} correspond to results for the \colorbox{Gainsboro}{N} top TD pairs. The same convention is used for \colorbox{Gainsboro}{literature} evidence and the \colorbox{Gainsboro}{time\_\_transition\_\_eq\_X} convention denotes RR estimates when the time since the phase 2 transition is equal to \colorbox{Gainsboro}{X} years. All other features are assessed based on their existence. The counts along the origin indicate how many TD pairs were used to compute the RR numerator.
  }
  \label{fig:relative_risk_core_features}
\end{figure}

@@

The primary model for this analysis, RDG, is a constrained, L2-regularized linear regressor. Alternatives to this choice we explore include constrained and unconstrained gradient-boosted machine (GBM) models fit with the LightGBM \cite{LightGBM} algorithm as well as unconstrained L2-regularized linear regressors (i.e. RDG with no constraints). The performance of these alternatives, as quantified by relative risk estimates among top predictions, is illustrated in Figure~\ref{fig:performance_metric_mean_across_ta}.

@@

\begin{figure}[!htb]
  \centering
  \captionsetup{width=.9\linewidth}
  \includegraphics[width=1\textwidth]{opportunity_summary.png}
  \caption{
    \textbf{Present-day target-disease pair counts by stage, likelihood of advancement and tractability}. The \textbf{stage} panel contains counts by maximum trial phase reached, the \textbf{threshold} panel contains counts of pairs with a RDG model score exceeding that of the associated benchmark for only undeveloped pairs, and the \textbf{tractability} panel shows pair frequencies among undeveloped pairs exceeding the EVA threshold that also have a HIGH tractability rating as defined in Supplementary~Table~\ref{tab:tractability_buckets}. This corresponds to targets that have all been in clinical development already, except for the \textbf{OC} modality in which case it indicates that a target has been approved. 
  }
  \label{fig:opportunity_summary}
\end{figure}

@@

@@

@@

\begin{figure}
  \centering
  \captionsetup{width=.9\linewidth}
  \includegraphics[width=1\textwidth]{relative_risk_model_features.pdf}
  \caption{
    \textbf{Performance by algorithm, constraint type and feature grouping}.  
  }
  \label{fig:relative_risk_model_features}
\end{figure}
